CrypTool Script\footnote{ Background reading for the CrypTool program}\\ Mathematics and Cryptography

### 2003-05-23 script/en/chapters/numbertheory.tex
be update

Number theory or arithmetic (the emphasis here is more on the aspect of performing calculations with numbers) 
was established by Carl Friedrich Gauss as a special mathematical discipline. Its elementary features 
include the greatest common divisor\footnote{This article deals with the gcd (greatest common 
divisor)\index{Gcd} in \hyperlink{Appendix_A}{Appendix A of this chapter}.} (gcd), congruence (remainder classes), 
factorisation, the Euler-Fermat theorem and primitive roots. However, the most important 
aspect is prime numbers and their multiplicative operation.

@@

\begin{center}
\fbox{\parbox{15cm}{{\em Carl Friedrich Gauss\index{Gauss} (30.4.1777$-$23.2.1855):}\\
Mathematics is the queen of sciences and number theory is the queen of mathematics.}}
\end{center}

@@

Number theory arose from interest in positive whole numbers $1, 2, 3, 4, \cdots$, also referred to as the set of natural numbers 
\index{Number!natural} {\em natural numbers} $\mathbb{N}$. These are the first mathematical constructs 
used by human civilisation. According to Kronecker\footnote{Leopold
  Kronecker, German mathematician, 1823$-$1891.}\index{Kronecker Leopold}, they are a creation of God. 
In Dedekind's\footnote{Julius Wilhelm Richard Dedekind,
  German mathematician, 06.10.1831$-$12.02.1916.}\index{Dedekind Julius} opinion, they are a creation of the human intellect. Dependent upon one's ideology,
this is an unsolvable contradiction or one and the same thing.

@@

Number theory has always been a source of fascination -- for both amateurs and professional mathematicians. In 
contrast to other areas of mathematics, many of the problems and theorems in number theory can be understood 
by non-experts. On the other hand, mathematicians often take a long time to find solutions to the problems or 
prove the theorems. It is therefore one thing to pose good questions but quite another matter to find the answer. One 
example of this is what is known as Fermat's Last (or large) theorem\footnote{One of the things you learn in mathematics at school is Pythagoras theorem, which states the following for a right-angle 
triangle: $a^2 + b^2 = c^2$, where $a$ and $b$ are the lengths of the sides containing the right angle and $c$ is the 
length of the hypotenuse. Fermat famously proposed that $a^n + b^n \not= c^n$ for whole-number exponents $n 
> 2$. Unfortunately, the letter in which Fermat made the claim did not have enough space for him to prove it. 
The theorem was not proved until over 300 years later \cite[p. 433-551]{Wiles1994}\index{Wiles}. }.

@@

\begin{center}
\fbox{\parbox{15cm}{{\em Joanne\index{Rowling} K. Rowling\footnotemark\:}\newline This isn't magic -- it's logic -- a puzzle. A 
lot of the greatest wizards haven't got an ounce of logic.}}
\end{center}

@@

This is the claim of the 1st {\em fundamental theorem of number theory} 
(= fundamental theorem of arithmetic = fundamental building block of all positive integers). This was 
formulated precisely for the first time by Carl Friedrich Gauss in his Disquisitiones Arithmeticae (1801).  
\index{Number theory!fundamental theorem}  \index{Gauss}

@@

\begin{center}
\fbox{\parbox{15cm}{{\em Eric Berne\footnotemark:\index{Berne}}\newline
Mathematical game theory postulates players who respond rationally.
Transactional game theory, on the other hand, deals with games that are 
not rational, perhaps even {\bf irrational and thereby closer to reality}.}}
\end{center}
\addtocounter{footnote}{0}\footnotetext{Eric Berne, ``Games People Play'', 
     rororo, (c) 1964, page 235.}
\vskip +4pt

@@

% --------------------------------------------------------------------------
\subsubsection{The Euler function\index{Euler!(phi) function}}
Given $n$, the number of numbers from the set $\{1, \cdots, n-1\}$ that are relatively prime to $n$ is equal to the value of 
the Euler\footnote{Leonhard Euler, Swiss mathematician, 15.4.1707 -- 18.9.1783\index{Euler}}
 function $J(n)$.

@@

% --------------------------------------------------------------------------
\subsubsection{The theorem of Euler-Fermat}\index{Euler}\index{RSA}
In order to prove the RSA procedure, we need Fermat's theorem and its 
generalisation (Euler-Fermat theorem) -- 
\hyperlink{KleinerSatzFermat-chap2}{please also see chapter 2.5}.

\hypertarget{KleinerSatzFermat-chap3}{}
\begin{theorem}\label{thm-zth-fermat1}{\bf Fermat's Little Theorem}\footnote{Pierre de Fermat, French mathematician, 17.8.1601 -- 12.1.1665.\index{Fermat}}
Let $p$ be a prime number and $a$ be a random whole number, then:
$$  a^p \equiv a {\rm ~(mod~} p).$$ 
\end{theorem}
An alternative formulation of Fermat's Little Theorem is as follows:
Let $p$ be a prime number and $a$ be a random whole number that is relatively prime to $p$, then:
$$      a^{p-1} \equiv 1 {\rm ~(mod~} p).$$

\begin{theorem}\label{thm-zth-fermateuler}{\bf Euler-Fermat theorem (generalisation of Fermat's Little Theorem)}
For all elements $a$ in the group $\mathbb{Z}_n^*$ (i.e. $a$ and $n$ are natural numbers that are co-prime):
$$a^{J(n)} \equiv 1 {\rm ~(mod~} n). $$
\end{theorem}

@@

The 2nd requirement can easily be met because there is a ``very'' large number of prime
numbers\footnote{According to the prime number theorem of Legendre and Gauss there are approximately $n/\ln(n)$ 
prime numbers\index{Prime number!number of} up to the number $n$. This means, for example, 
that there are $6.5*10^{74}$ prime numbers under $n=2^{256}$ ($=1.1*10^{77}$) and $3.2*10^{74}$ prime numbers under $n=2^{255}$. 
Between $2^{255}$ and $2^{256}$ there are therefore $3.3*10^{74}$ prime numbers with precisely $256$ bits. This large number is 
also the reason why we cannot simply save them all.} and because this can be ensured by a central office that issues certificates.

@@

This formula shows, that the factorisation problem belongs to the class of
problems with sub-exponential time complexity (i.~e.\ time complexity grows
asymptotically not as fast as exponential functions like $e^l$ or $2^l$,
but strictly slower, e.~g.\ like $e^{\sqrt{l}}$).  This classification is
all that is currently known; it does not preclude the possibility that the
factorisation problem can be solved in polynomial time (see~\ref{RSABernstein}).

@@

In August 2002 the three Indian researchers M. Agrawal, N. Kayal and N. Saxena published the paper {\em ``PRIMES in P''} about a new primality testing algorithm 
\cite{Agrawal2002}. 
They discovered a polynomial time deterministic algorithm for determining if a number is prime or not.

@@

The current record (as of May 2003) obtained using the GNFS method
(General Number Field Sieve) factorised a 160-digit into its both prime
factors (these are build with each 80 decimal digits).

@@

On January 18, 2002 researchers of the German University of Bonn\footnote{%
   \href{http://www.loria.fr/~zimmerma/records/rsa160}
{\texttt{http://www.loria.fr/~zimmerma/records/rsa160}}} 
factorised a 160-digit number into its both prime factors (these are build
with each 80 decimal digits) using the GNFS method (General Number Field
Sieve).

@@

-----------
The composite number called ``RSA-160'' is (the following 3 lines contain
one number):
$$
\begin{array}{c}
215274110271888970189601520131282542925777358884567598017049 \\
767677813314521885913567301105977349105960249790711158521430 \\
2079314665202840140619946994927570407753
\end{array}
$$
The factorisation of RSA-160 resulted in the following two prime factors:
$$
\begin{array}{c}
p = 45427892858481394071686190649738831 \\         
    656137145778469793250959984709250004157335359
\end{array}
$$
and
$$
\begin{array}{c}
q = 47388090603832016196633832303788951 \\
    973268922921040957944741354648812028493909367
\end{array}
$$

@@

\begin{minipage}{\textwidth}
\vspace{3ex}
Links:
\vspace{-10pt}
\begin{itemize}
\item[]   \href{http://www.loria.fr/~zimmerma/records/rsa160}
       {\texttt{http://www.loria.fr/~zimmerma/records/rsa160}} \\
          \href{http://www.loria.fr/~zimmerma/records/factor.html}
       {\texttt{http://www.loria.fr/~zimmerma/records/factor.html}} \\
          \href{http://www.crypto-world.com/FactorWorld.html}
       {\texttt{http://www.crypto-world.com/FactorWorld.html}}
\end{itemize}
\end{minipage}
\vspace{12pt}

@@

\begin{center}
\fbox{\parbox{15cm}{{\em Joanne\index{Rowling} K. Rowling\footnotemark:}\\
It is our choices, that show what we truly are, far more than our abilities.}}
\end{center}

@@

% --------------------------------------------------------------------------
\subsubsection[RSA with slightly larger primes]{RSA with slightly larger primes and a text of upper case letters}
\label{rsaex2}

@@

\bibitem[Wiles1994]{Wiles1994} \index{Wiles}
    Wiles, Andrew, \\
    {\em Modular elliptic curves and Fermat's Last Theorem}, \\
    \index{Fermat!last theorem}
    in Annals of Mathematics 141 (1995).

@@

   \item \index{Munchenbach@M\"unchenbach}
         Only after I had completed this article, did I come across the
         website of Mr.\ M\"unchenbach, which interactively and didactically
         uses elementary number theory to provide a sophisticated
         description of the fundamental mathematical thought processes. 
         It was created for a teaching project in the 11th grade 
         of the technical grammar school (unfortunately only available in German): \\
         \href{http://www.hydrargyrum.de/kryptographie}
         {\texttt{http://www.hydrargyrum.de/kryptographie}}

@@

   \item Factorisation records and challenges \index{Factorisation!factorisation records}\\
         \href{http://www.crypto-world.com/}
         {\texttt{http://www.crypto-world.com/}} \\
         \href{http://www.crypto-world.com/FactorWorld.html}
         {\texttt{http://www.crypto-world.com/FactorWorld.html, page by Scott Contini}} \\
         \href{http://www.loria.fr/~zimmerma/records/factor.html}
         {\texttt{http://www.loria.fr/~zimmerma/records/factor.html}} \\

         \href{http://www.uni-bonn.de/Aktuelles/Pressemitteilungen/pm02/pm035-02.html}
         {\texttt{http://www.uni-bonn.de/Aktuelles/Pressemitteilungen/pm02/pm035-02.html}} \\
         \href{http://www.ercim.org/publication/Ercim\_News/enw49/franke.html, 2002-01}
         {\texttt{http://www.ercim.org/publication/Ercim\_News/enw49/franke.html, 2002-01}} \\
         \href{http://www.loria.fr/~zimmerma/records/rsa160}
         {\texttt{http://www.loria.fr/~zimmerma/records/rsa160}} \\

>>>>

Number theory or arithmetic (the emphasis here is more on the aspect of 
performing calculations with numbers) was established by Carl Friedrich Gauss
\index{Gauss Carl Friedrich} as a special mathematical discipline. Its 
elementary features include the greatest common divisor\footnote{This 
article deals with the gcd (greatest common 
divisor)\index{Gcd} in \hyperlink{Appendix_A}{Appendix A of this chapter}.}
(gcd), congruence (remainder classes), factorisation, the Euler-Fermat theorem
and primitive roots. However, the most important 
aspect is prime numbers and their multiplicative operation.

@@

\begin{center}
\fbox{\parbox{15cm}{{\em Carl Friedrich Gauss\index{Gauss Carl Friedrich}
(Apr 4, 1777$-$Feb 23, 1855):}\\
Mathematics is the queen of sciences and number theory is the queen of mathematics.}}
\end{center}

@@

Number theory arose from interest in positive whole numbers $1, 2, 3, 4, \cdots$, also referred to as the set of natural numbers 
\index{Number!natural} {\em natural numbers} $\mathbb{N}$. These are the first mathematical constructs 
used by human civilisation. According to Kronecker\footnote{Leopold
  Kronecker, German mathematician, Dec 7, 1823$-$Dec 29, 1891}\index{Kronecker Leopold}, they are a creation of God. 
In Dedekind's\footnote{Julius Wilhelm Richard Dedekind,
  German mathematician, 06.10.1831$-$12.02.1916.}\index{Dedekind Julius} opinion, they are a creation of the human intellect. Dependent upon one's ideology,
this is an unsolvable contradiction or one and the same thing.

@@

Number theory has always been a source of fascination -- for both amateurs and professional mathematicians. In 
contrast to other areas of mathematics, many of the problems and theorems in number theory can be understood 
by non-experts. On the other hand, mathematicians often take a long time to find solutions to the problems or 
prove the theorems. It is therefore one thing to pose good questions but quite another matter to find the answer. One 
example of this is what is known as Fermat's Last (or large) theorem\footnote{One of the things you learn in mathematics at school is Pythagoras theorem, which states the following for a right-angle 
triangle: $a^2 + b^2 = c^2$, where $a$ and $b$ are the lengths of the sides containing the right angle and $c$ is the 
length of the hypotenuse. Fermat famously proposed that $a^n + b^n \not= c^n$ for whole-number exponents $n 
> 2$. Unfortunately, the letter in which Fermat made the claim did not have enough space for him to prove it. 
The theorem was not proved until over 300 years later \cite[p. 433-551]{Wiles1994}\index{Wiles Andrew}. }.

@@

\begin{center}
\fbox{\parbox{15cm}{{\em Joanne\index{Rowling Joanne} K. Rowling\footnotemark\:}\newline This isn't magic -- it's logic -- a puzzle. A 
lot of the greatest wizards haven't got an ounce of logic.}}
\end{center}

@@

This is the claim of the 1st {\em fundamental theorem of number theory} 
(= fundamental theorem of arithmetic = fundamental building block of all positive integers). This was 
formulated precisely for the first time by Carl Friedrich Gauss in his Disquisitiones Arithmeticae (1801).  
\index{Number theory!fundamental theorem}  \index{Gauss Carl Friedrich}

@@

\begin{center}
\fbox{\parbox{15cm}{{\em Eric Berne\footnotemark:\index{Berne Eric}}\newline
Mathematical game theory postulates players who respond rationally.
Transactional game theory, on the other hand, deals with games that are 
not rational, perhaps even {\bf irrational and thereby closer to reality}.}}
\end{center}
\addtocounter{footnote}{0}\footnotetext{Eric Berne, ``Games People Play'', 
     rororo, (c) 1964, page 235.}
\vskip +4pt

@@

% --------------------------------------------------------------------------
\subsubsection{The Euler function\index{Euler!(phi) function}}
Given $n$, the number of numbers from the set $\{1, \cdots, n-1\}$ that are relatively prime to $n$ is equal to the value of 
the Euler\footnote{Leonhard Euler, Swiss mathematician, Apr 15, 1707 -- Sep 18, 1783\index{Euler Leonhard}}
 function $J(n)$.

@@

% --------------------------------------------------------------------------
\subsubsection{The theorem of Euler-Fermat}\index{Euler Leonhard}\index{RSA}
In order to prove the RSA procedure, we need Fermat's theorem and its 
generalisation (Euler-Fermat theorem) -- 
\hyperlink{KleinerSatzFermat-chap2}{please also see chapter 2.5}.

\hypertarget{KleinerSatzFermat-chap3}{}
\begin{theorem}\label{thm-zth-fermat1}{\bf Fermat's Little Theorem}\footnote{Pierre de Fermat, French mathematician, Aug 17, 1601 -- Jan 12, 1665.
\index{Fermat Pierre}
}
Let $p$ be a prime number and $a$ be a random whole number, then:
$$  a^p \equiv a {\rm ~(mod~} p).$$ 
\end{theorem}
An alternative formulation of Fermat's Little Theorem is as follows:
Let $p$ be a prime number and $a$ be a random whole number that is relatively prime to $p$, then:
$$      a^{p-1} \equiv 1 {\rm ~(mod~} p).$$

\begin{theorem}\label{thm-zth-fermateuler}{\bf Euler-Fermat theorem (generalisation of Fermat's Little Theorem)}
For all elements $a$ in the group $\mathbb{Z}_n^*$ (i.e. $a$ and $n$ are natural numbers that are co-prime):
$$a^{J(n)} \equiv 1 {\rm ~(mod~} n). $$
\end{theorem}
\index{Euler Leonhard} \index{Fermat Pierre}

@@

The 2nd requirement can easily be met because there is a ``very'' large number
of prime numbers\footnote{%
According to the prime number theorem of 
Legendre\index{Legendre Adrien-Marie} and Gauss \index{Gauss Carl Friedrich} 
there are approximately $n/\ln(n)$ 
prime numbers\index{Prime number!number of} up to the number $n$. 
This means, for example, that there are $6.5*10^{74}$ prime numbers 
under $n=2^{256}$ ($=1.1*10^{77}$) and $3.2*10^{74}$ prime numbers 
under $n=2^{255}$. 
Between $2^{255}$ and $2^{256}$ there are therefore $3.3*10^{74}$ prime numbers with precisely $256$ bits. This large number is 
also the reason why we cannot simply save them all.} and because this can be ensured by a central office that issues certificates.

@@

This formula shows, that the factorisation problem belongs to the class of
problems with sub-exponential time complexity (i.~e.\ time complexity grows
asymptotically not as fast as exponential functions like $e^l$ or $2^l$,
but strictly slower, e.~g.\ like $e^{\sqrt{l}}$).  This classification is
all that is currently known; it does not preclude the possibility that the
factorisation problem can be solved in polynomial\index{Polynomial} time 
(see~\ref{RSABernstein}).

@@

In August 2002 the three Indian researchers M. Agrawal, N. Kayal and N. Saxena published the paper {\em ``PRIMES in P''} about a new primality testing algorithm called AKS\index{AKS} \cite{Agrawal2002}. 
They discovered a polynomial\index{Polynomial} time deterministic algorithm for determining if a number is prime or not.

@@

The current record (as of May 2003) obtained using the GNFS method
(General Number Field Sieve) factorised a 160-digit into its both prime
factors.

@@

On January 18, 2002 researchers of the German University of Bonn\footnote{%
   \href{http://www.loria.fr/~zimmerma/records/rsa160}
{\texttt{http://www.loria.fr/\~{}zimmerma/records/rsa160}}} 
factorised a 160-digit number into its both prime factors (these are build
with each 80 decimal digits) using the GNFS method (General Number Field
Sieve).

@@

The composite number called ``RSA-160'' is (the following 3 lines contain
one number):
$$
\begin{array}{c}
215274110271888970189601520131282542925777358884567598017049 \\
767677813314521885913567301105977349105960249790711158521430 \\
2079314665202840140619946994927570407753
\end{array}
$$
The factorisation of RSA-160 resulted in the following two prime factors:
$$
\begin{array}{c}
p = 45427892858481394071686190649738831 \\         
    656137145778469793250959984709250004157335359
\end{array}
$$
and
$$
\begin{array}{c}
q = 47388090603832016196633832303788951 \\
    973268922921040957944741354648812028493909367
\end{array}
$$

@@

\begin{minipage}{\textwidth}
\vspace{3ex}
Links:
\vspace{-10pt}
\begin{itemize}
\item[]   \href{http://www.loria.fr/~zimmerma/records/rsa160}
       {\texttt{http://www.loria.fr/\~{}zimmerma/records/rsa160}} \\
          \href{http://www.loria.fr/~zimmerma/records/factor.html}
       {\texttt{http://www.loria.fr/\~{}zimmerma/records/factor.html}} \\
          \href{http://www.crypto-world.com/FactorWorld.html}
       {\texttt{http://www.crypto-world.com/FactorWorld.html}}
\end{itemize}
\end{minipage}
\vspace{12pt}

\vspace{24pt}

@@

\begin{center}
\fbox{\parbox{15cm}{{\em Joanne\index{Rowling Joanne} K. Rowling\footnotemark:}\\
It is our choices, that show what we truly are, far more than our abilities.}}
\end{center}

@@

% --------------------------------------------------------------------------
\subsubsection[RSA with slightly larger primes and an upper-case message]{RSA with slightly larger primes and a text of upper case letters}
\label{rsaex2}

@@

\bibitem[Wiles1994]{Wiles1994} \index{Wiles Andrew}
    Wiles, Andrew, \\
    {\em Modular elliptic curves and Fermat's Last Theorem}, \\
    \index{Fermat!last theorem}
    in Annals of Mathematics 141 (1995).

@@

   \item \index{Munchenbach@M\"unchenbach Carsten}
         Only after I had completed this article, did I come across the
         website of Mr.\ M\"unchenbach, which interactively and didactically
         uses elementary number theory to provide a sophisticated
         description of the fundamental mathematical thought processes. 
         It was created for a teaching project in the 11th grade 
         of the technical grammar school (unfortunately only available in German): \\
         \href{http://www.hydrargyrum.de/kryptographie}
         {\texttt{http://www.hydrargyrum.de/kryptographie}}

@@

   \item Factorisation records and challenges\index{Factorisation!factorisation records},\\
         \href{http://www.crypto-world.com/}
         {\texttt{http://www.crypto-world.com/}} \\
         \href{http://www.crypto-world.com/FactorWorld.html}
         {\texttt{http://www.crypto-world.com/FactorWorld.html, page by Scott Contini}} \\
         \href{http://www.loria.fr/~zimmerma/records/factor.html}
         {\texttt{http://www.loria.fr/\~{}zimmerma/records/factor.html}} \\

         \href{http://www.uni-bonn.de/Aktuelles/Pressemitteilungen/pm02/pm035-02.html}
         {\texttt{http://www.uni-bonn.de/Aktuelles/Pressemitteilungen/pm02/pm035-02.html}} \\
         \href{http://www.ercim.org/publication/Ercim\_News/enw49/franke.html, 2002-01}
         {\texttt{http://www.ercim.org/publication/Ercim\_News/enw49/franke.html, 2002-01}} \\
         \href{http://www.loria.fr/~zimmerma/records/rsa160}
         {\texttt{http://www.loria.fr/\~{}zimmerma/records/rsa160}} \\

### 2003-05-23 script/en/chapters/primes.tex
be update

% $Id$
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% C H A P T E R   T W O
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\hypertarget{FermatNumbers01}{}
In the 17th century, Fermat wrote to Mersenne \index{Mersenne Marin} that he presumed that all numbers
of the form $$ F(n) = 2^{2^n} + 1 $$ are prime for all whole numbers $n$ greater
than or equal to $0$ (\hyperlink{FermatNumbers02}{see below}) 
\index{Number!Fermat}\index{Fermat!number}

@@

A stronger test is provided by\index{Miller}\index{Rabin} Miller/Rabin: it is
only passed by so-called {\em strong pseudo prime numbers}. Again, there are
strong pseudo prime numbers that are not primes, but this is much less often the
case than for (simple) pseudo prime numbers. The smallest strong pseudo prime
number base $2$ is $$ 15,841 = 7 \cdot 31 \cdot 73. $$
If you test all 4 bases, $2, 3, 5$ and $7$, you will find only one strong
pseudo prime number up to $25 \cdot 10^9$, i.e. a number that passes the 
test and yet is not a prime number.

@@

% --------------------------------------------------------------------------
\vskip +20 pt
\subsection{Further special number types and the search for a formula for primes}
\index{Prime number!formula}
There are currently no useful, open (i.e. not recursive) formulae known that
only deliver prime numbers (recursive means that in order to calculate the
function the same function is used with a smaller variable). Mathematicians
would be happy if they could find a formula that leaves gaps (i.e. does not
deliver all prime numbers) but does not deliver any composite (non-prime)
numbers.

@@

\vskip +15 pt
\hypertarget{FermatNumbers02}{}
\item {\em Fermat numbers}\footnote{The Fermat prime numbers play a role in circle division.
As proven by Gauss\index{Gauss}, a regular $p$-edge
can only be constructed with the use of a pair of compasses and a ruler, when $p$ is a Fermat prime number.}
\index{Number!Fermat}\index{Fermat!number}  $F(n) = 2^{2^n} + 1$:\\
As mentioned \hyperlink{FermatNumbers01}{above}, Fermat wrote to Mersenne regarding his assumption, that all numbers of this type are primes.
Surprisingly he would have been able obtain a positive result using the negative
prime number test for $n=5$ based on his small theorem.
$$
\begin{array}{lll}
F(0) = 2^{2^0} + 1  = 2^1 + 1 & = 3 &   \mapsto {\rm ~prime}  \\
F(1) = 2^{2^1} + 1  = 2^2 + 1 & = 5 &   \mapsto {\rm ~prime}  \\
F(2) = 2^{2^2} + 1  = 2^4 + 1 & = 17 &  \mapsto {\rm ~prime}  \\
F(3) = 2^{2^3} + 1  = 2^8 + 1 & = 257 & \mapsto {\rm ~prime}  \\
F(4) = 2^{2^4} + 1  = 2^{16} + 1 &  = \mbox{65,537} &  \mapsto {\rm ~prime}  \\
F(5) = 2^{2^5} + 1  = 2^{32} + 1 &  = \mbox{4,294,967,297} = 641 \cdot \mbox{6,700,417} &  \mapsto {\rm ~NOT~prime} ! \\
F(6) = 2^{2^6} + 1  = 2^{64} + 1 &  = \mbox{18,446,744,073,709,551,617} \\
                                 &  = \mbox{274,177} \cdot \mbox{67,280,421,310,721} & \mapsto {\rm ~NOT~prime} !
\end{array} 
$$

@@

\newpage
\item Polynomial functions $f(x) = a_n x^n + a_{n-1}x^{n-1} + \cdots + a_1 x^1 +
a_0$  ($a_i$ in ${\mathbb Z}$, $n \geq 1$):

@@

  \item  \index{Catalan number} Catalan, after whom the so-called
{\em Catalan numbers} $A(n) = (1 / (n+1) ) * (2n)! / (n!)^2$ are named,
conjectured that $C_4$ is a prime:
$$
 \begin{array}{l}
C_0 = 2, \\
C_1 = 2^{C_0} - 1,  \\
C_2 = 2^{C_1} - 1,  \\
C_3 = 2^{C_2} - 1, \\
C_4 = 2^{C_3} - 1, \cdots \\
\end{array}
$$
    (see {\href{http://www.utm.edu/research/primes/mersenne.shtml}{\tt
http://www.utm.edu/research/primes/mersenne.shtml}} under Conjectures and
Unsolved Problems).

@@

This is the \index{Prime number!theorem} \textbf{prime number theorem}. It was
put forward by Legendre \index{Legendre} (1752-1833) and Gauss\index{Gauss} (1777-1855) but not proved until
over 100 years later.

@@

  \item Does a polynomial time algorithm exist for calculating the prime factors of a number (see
    \cite[p. 167]{Klee1997})? 
    This question can be divided into the two following questions:
\begin{itemize} 
 \item Does a polynomial time algorithm exist that decides whether a number is
prime?
\item Does a polynomial time algorithm exist that calculates for a
  composite number $n$ a non-trivial (i.e. other than $1$ and $n$) factor
  of $n$?\footnote{Please see chapters \ref{RSABernstein} and
    \ref{NoteFactorisation}.}
\end{itemize}
\end{itemize}

@@

\item The best overview about prime numbers is offered from my point of view 
      by ~``The Prime Pages'' from Chris Caldwell.
      \index{Chris Caldwell} \\
      \href{http://www.utm.edu/research/primes}
       {\tt http://www.utm.edu/research/primes }

>>>>

% $Id$
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% C H A P T E R   T W O:  P R I M E S
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\hypertarget{FermatNumbers01}{}%
In the 17th century, Fermat\index{Fermat Pierre} wrote to Mersenne \index{Mersenne Marin} that he presumed that all numbers
of the form $$ F(n) = 2^{2^n} + 1 $$ are prime for all whole numbers $n$ greater
than or equal to $0$ (\hyperlink{FermatNumbers02}{see below}) 
\index{Number!Fermat}\index{Fermat!number}

@@

A stronger test is provided by\index{Miller G.L.}\index{Rabin} Miller/Rabin: it is
only passed by so-called {\em strong pseudo prime numbers}. Again, there are
strong pseudo prime numbers that are not primes, but this is much less often the
case than for (simple) pseudo prime numbers. The smallest strong pseudo prime
number base $2$ is $$ 15,841 = 7 \cdot 31 \cdot 73. $$
If you test all 4 bases, $2, 3, 5$ and $7$, you will find only one strong
pseudo prime number up to $25 \cdot 10^9$, i.e. a number that passes the 
test and yet is not a prime number.

@@

% --------------------------------------------------------------------------
\vskip +20 pt
\subsection{Further special number types and the search for a formula for primes}\label{spezialzahlentypen}
\index{Prime number!formula}
There are currently no useful, open (i.e. not recursive) formulae known that
only deliver prime numbers (recursive means that in order to calculate the
function the same function is used with a smaller variable). Mathematicians
would be happy if they could find a formula that leaves gaps (i.e. does not
deliver all prime numbers) but does not deliver any composite (non-prime)
numbers.

@@

\vskip +15 pt
\hypertarget{FermatNumbers02}{}
\item {\em Fermat numbers}\footnote{The Fermat prime numbers play a role in circle division.
As proven by Gauss\index{Gauss Carl Friedrich}, a regular $p$-edge
can only be constructed with the use of a pair of compasses and a ruler, when $p$ is a Fermat prime number.}
\index{Number!Fermat}\index{Fermat!number}  $F(n) = 2^{2^n} + 1$:\\
As mentioned \hyperlink{FermatNumbers01}{above}, Fermat wrote to Mersenne regarding his assumption, that all numbers of this type are primes.
Surprisingly he would have been able obtain a positive result using the negative
prime number test for $n=5$ based on his small theorem.
$$
\begin{array}{lll}
F(0) = 2^{2^0} + 1  = 2^1 + 1 & = 3 &   \mapsto {\rm ~prime}  \\
F(1) = 2^{2^1} + 1  = 2^2 + 1 & = 5 &   \mapsto {\rm ~prime}  \\
F(2) = 2^{2^2} + 1  = 2^4 + 1 & = 17 &  \mapsto {\rm ~prime}  \\
F(3) = 2^{2^3} + 1  = 2^8 + 1 & = 257 & \mapsto {\rm ~prime}  \\
F(4) = 2^{2^4} + 1  = 2^{16} + 1 &  = \mbox{65,537} &  \mapsto {\rm ~prime}  \\
F(5) = 2^{2^5} + 1  = 2^{32} + 1 &  = \mbox{4,294,967,297} = 641 \cdot \mbox{6,700,417} &  \mapsto {\rm ~NOT~prime} ! \\
F(6) = 2^{2^6} + 1  = 2^{64} + 1 &  = \mbox{18,446,744,073,709,551,617} \\
                                 &  = \mbox{274,177} \cdot \mbox{67,280,421,310,721} & \mapsto {\rm ~NOT~prime} !
\end{array} 
$$

@@

\newpage
\item Polynomial functions\index{Polynomial} $f(x) = a_n x^n + a_{n-1}x^{n-1} + \cdots + a_1 x^1 +
a_0$  ($a_i$ in ${\mathbb Z}$, $n \geq 1$):

@@

  \item \index{Catalan Eugene} \index{Catalan number} Eugene Catalan\footnote{%
Belgian mathematician, May 5, 1814 - Feb 14, 1894.

After him the so-called
{\em Catalan numbers}
$A(n) = (1 / (n+1) ) * (2n)! / (n!)^2  =  1, 2, 5, 14, 42, 132, 429, 1430, 
4862, 16796, 58786, 208012, 742900, 2674440, 9694845, ... $
are named.}
conjectured that $C_4$ is a prime:
$$
 \begin{array}{l}
C_0 = 2, \\
C_1 = 2^{C_0} - 1,  \\
C_2 = 2^{C_1} - 1,  \\
C_3 = 2^{C_2} - 1, \\
C_4 = 2^{C_3} - 1, \cdots \\
\end{array}
$$
    (see {\href{http://www.utm.edu/research/primes/mersenne.shtml}{\tt
http://www.utm.edu/research/primes/mersenne.shtml}} under Conjectures and
Unsolved Problems).

@@

This is the \index{Prime number!theorem} \textbf{prime number theorem}. 
It was put forward by Legendre\footnote{%
Adrien-Marie Legendre, French mathematician, Sep 18, 1752$-$Jan 10, 1833.
}
\index{Legendre Adrien-Marie}
and Gauss\index{Gauss Carl Friedrich} (1777-1855) but not proved until
over 100 years later.

@@

  \item Does a polynomial\index{Polynomial} time algorithm exist for calculating the prime factors of a number (see
    \cite[p. 167]{Klee1997})? 
    This question can be divided into the two following questions:
   \begin{itemize} 
      \item Does a polynomial time algorithm exist that decides whether a
            number is prime?
	    This question has been answered by the AKS algorithm\index{AKS} 
	    (see chapter\ref{PrimesinP}, ``Primes in P'': 
	    Primality testing is polynomial).

      \item Does a polynomial time algorithm exist that calculates for a
            composite number $n$ a non-trivial (i.e. other than $1$ and $n$)
	    factor of $n$?\footnote{Please see chapters \ref{RSABernstein}
	    and \ref{NoteFactorisation}.}
   \end{itemize}
\end{itemize}

@@

\item The best overview about prime numbers is offered from my point of view 
      by ~``The Prime Pages'' from Chris Caldwell.
      \index{Caldwell Chris} \\
      \href{http://www.utm.edu/research/primes}
       {\tt http://www.utm.edu/research/primes }
